---
title: "Microbiome Power and Effect Size Analysis"
author: "Analysis Report"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
# Configure chunk options for a clean PDF output
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, 
  message = FALSE,
  collapse = TRUE,
  comment = ""
)
```

## Setup and Data Loading

First, we load the necessary libraries, set up the project directories, and import the phyloseq object.

```{r init}
# 1. Define standard CRAN packages
cran_packages <- c("pwr", "dplyr", "vegan")

# Install missing CRAN packages
for (pkg in cran_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    install.packages(pkg)
  }
}

# 2. Install Bioconductor package (phyloseq) if missing
if (!requireNamespace("phyloseq", quietly = TRUE)) {
  if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
  }
  # update = FALSE prevents interactive prompts that can crash R Markdown renders
  BiocManager::install("phyloseq", update = FALSE, ask = FALSE) 
}

# 3. Load all packages
library(pwr)
library(dplyr)
library(phyloseq)
library(vegan)

# Setup directories
# Note: if this.path::here() fails during knitting, you may need to 
# replace it with here::here() or an absolute path.
project_dir <- this.path::here(..=1)
results_dir <- file.path(project_dir, "results/alpha_diversity")
dir.create(results_dir, showWarnings = FALSE)

in_dir <- file.path(project_dir, "data/phyloseq")
ps <- readRDS(file = file.path(in_dir, "/phyloseq_object_species.rds"))

# Pre-processing: remove unknown taxa and factorize subgroup
ps <- prune_taxa(taxa_names(ps) != "Unknown", ps)
ps@sam_data$subgroup <- factor(
  ps@sam_data$subgroup,
  levels = c("tooth_12m", "tooth_baseline", "implant_baseline", "implant_12m")
)
```

## Rarefaction and Alpha Diversity Metrics

We rarefy the data to an even depth to standardize sequencing effort across samples, then compute three alpha diversity measures: Observed, Shannon, and Simpson.

```{r alpha_calc}
set.seed(1)
ps.rarefied <- rarefy_even_depth(
  ps, 
  rngseed = 1, 
  sample.size = min(sample_sums(ps)), 
  replace = FALSE
)

alpha <- estimate_richness(
  ps.rarefied, 
  split = TRUE,
  measures = c("Observed", "Shannon", "Simpson")
)

alpha$sample_id <- rownames(alpha)
alpha <- merge(
  alpha,
  data.frame(sample_data(ps.rarefied)),
  by.x = "sample_id", 
  by.y = "row.names"
)

alpha$subgroup <- factor(alpha$subgroup)  # 4 levels
```

## Alpha Diversity Power Analysis

Here, we define a function to compute both the p-value and Cohen's $f$ effect size using a one-way ANOVA. 

```{r alpha_power}
# Helper: compute Cohen's f and extract p-value from one-way ANOVA
effect_size_f <- function(y, group){
  df <- data.frame(y = y, group = group)
  df <- df[complete.cases(df), ]
  aov_fit <- aov(y ~ group, data = df)
  ss <- summary(aov_fit)[[1]]
  
  # Extract and print the ANOVA p-value
  p_val <- ss["group", "Pr(>F)"]
  cat("ANOVA p-value:", format.pval(p_val), "\n")
  
  SSb <- ss["group", "Sum Sq"]              # between-groups SS
  SSt <- sum(ss[,"Sum Sq"])                 # total SS
  eta2 <- SSb / SSt                         # eta-squared
  f <- sqrt(eta2 / (1 - eta2))              # Cohen's f
  return(f)
}

# Function to compute power-based N for one alpha metric
power_for_metric <- function(metric_name){
  y <- alpha[[metric_name]]
  
  cat("\n=========================================\n")
  cat("Metric:", metric_name, "\n")
  
  # Calculate effect size and print p-value (handled inside effect_size_f)
  f <- effect_size_f(y, alpha$subgroup)
  
  cat("Estimated effect size (Cohen's f):", round(f, 3), "\n")
  
  # If effect is ~0 (no signal), power calc is not meaningful
  if(is.na(f) || f <= 0){
    cat("Effect size is zero or undefined; cannot compute meaningful sample size.\n")
    return(invisible(NULL))
  }
  
  # Power analysis: 4 groups, 80% power, alpha = 0.05
  pwr_res <- pwr.anova.test(k = nlevels(alpha$subgroup),
                            f = f,
                            sig.level = 0.05,
                            power = 0.80)
  
  cat("Required total sample size for 80% power:", ceiling(pwr_res$n * pwr_res$k), "\n")
  cat("Required sample size per group (assuming equal n):", ceiling(pwr_res$n), "\n")
  
  invisible(pwr_res)
}

# Run for the three alpha-diversity metrics
res_observed <- power_for_metric("Observed")
res_shannon  <- power_for_metric("Shannon")
res_simpson  <- power_for_metric("Simpson")
```

## Beta Diversity Power Analysis

Next, we calculate the Bray-Curtis and Jaccard distance matrices and run PERMANOVA (`adonis2`). We will extract the exact p-value and $R^2$, then convert $R^2$ to Cohen's $f$ to calculate the required sample sizes.

```{r beta_power}
bray_dist <- phyloseq::distance(ps.rarefied, method = "bray")
jaccard_dist <- phyloseq::distance(ps.rarefied, method = "jaccard", binary = TRUE)

meta_df <- data.frame(sample_data(ps.rarefied))
meta_df <- meta_df[match(rownames(as.matrix(bray_dist)), rownames(meta_df)), ]

# PERMANOVA for Bray-Curtis and Jaccard
adonis_bray <- vegan::adonis2(as.matrix(bray_dist) ~ subgroup, data = meta_df)
adonis_jaccard <- vegan::adonis2(as.matrix(jaccard_dist) ~ subgroup, data = meta_df)

# Extract R² values
R2_bray <- adonis_bray[1, "R2"]
R2_jaccard <- adonis_jaccard[1, "R2"]

# Extract p-values
p_val_bray <- adonis_bray[1, "Pr(>F)"]
p_val_jaccard <- adonis_jaccard[1, "Pr(>F)"]

cat("\n=========================================\n")
cat("Beta Diversity Analysis\n")
cat("-----------------------------------------\n")
cat("Bray-Curtis:\n")
cat("  PERMANOVA p-value =", format.pval(p_val_bray), "\n")
cat("  R² =", round(R2_bray, 3), "\n")

cat("\nJaccard:\n")
cat("  PERMANOVA p-value =", format.pval(p_val_jaccard), "\n")
cat("  R² =", round(R2_jaccard, 3), "\n")

# Convert R² to Cohen's f
f_bray    <- sqrt(R2_bray    / (1 - R2_bray))
f_jaccard <- sqrt(R2_jaccard / (1 - R2_jaccard))

cat("\nEffect Sizes:\n")
cat("  Bray-Curtis f =", round(f_bray, 3), "\n")
cat("  Jaccard f =", round(f_jaccard, 3), "\n")

# Function to compute required N from f
beta_power_from_f <- function(f, k = 4, power = 0.80, alpha = 0.05){
  if(is.na(f) || f <= 0){
    cat("Effect size f is zero/NA; cannot compute meaningful sample size.\n")
    return(invisible(NULL))
  }
  res <- pwr.anova.test(k = k, f = f, sig.level = alpha, power = power)
  total_n <- ceiling(res$n * k)
  per_group <- ceiling(res$n)
  list(result = res, total_n = total_n, per_group = per_group)
}

cat("\n-----------------------------------------\n")
# Power for Bray-Curtis
pow_bray <- beta_power_from_f(f_bray, k = nlevels(meta_df$subgroup))
cat("Bray-Curtis PERMANOVA Power:\n")
cat("  Required total N for 80% power:", pow_bray$total_n, "\n")
cat("  Required N per group:", pow_bray$per_group, "\n\n")

# Power for Jaccard
pow_jac <- beta_power_from_f(f_jaccard, k = nlevels(meta_df$subgroup))
cat("Jaccard PERMANOVA Power:\n")
cat("  Required total N for 80% power:", pow_jac$total_n, "\n")
cat("  Required N per group:", pow_jac$per_group, "\n")
```